---
title: "2: Physical Properties of Lakes"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Investigate the concepts of lake stratification and mixing by analyzing monitoring data
2. Apply data analytics skills to applied questions about physical properties of lakes
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

What are the physical properties of lakes?
* Temperature
* Depth
* Light transmission

## Session Set Up
```{r, message = FALSE}
# Check working directory (should be project file location)
getwd()

# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("rLakeAnalyzer")

# load packages
library(tidyverse)
library(lubridate)
library(rLakeAnalyzer)

# Load data
NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv") #data from north temperate lakes ecological research station

# set ggplot theme
theme_set(theme_classic())
```

## Data Wrangling and Exploration

### Investigate structure of dataset

Common steps/checks for data wrangling: 

* Date formatting
* Addressing NAs
* Subsetting and filtering 

```{r}
# Is the date column perceived as a date by R?
class(NTLdata$sampledate) #ask R what kind of data it sees in the column "sampledate"
NTLdata$sampledate <- as.Date(NTLdata$sampledate, format = "%m/%d/%y") #tell R to read the class as a date and provide R with the format that the date exists in. y=2-digit year, Y=4-digit year. help(as.Date) 
class(NTLdata$sampledate)
# What does this column look like now?
    #now looks like YYY-MM-DD (universal format)

# Remove rows that contain NAs in temperature column
dim(NTLdata) #check dimensions of dataset...output: # of rows, # of columns
NTLdata <- NTLdata %>% #pipe "%>%" tells R to continue the function..."then"
  drop_na(temperature_C) #delete rows that don't have any temperature data 
dim(NTLdata)  
        #to write in one line...
              #NTLdata <- drop_na(NTLdata, temperature_C)
      

# How many observations are there for each lake?
summary(NTLdata$lakename) #R is treating the column as a character
summary(as.factor(NTLdata$lakename)) #count of how many rows in the dataset that each lake name appears in the dataset. also tells us where we have the most temperature data (peter lake has the most temperature data) bc we removed all the rows that do not have temperature data

# Let's choose the two lakes with the most data
NTLdata_PeterPaul <- NTLdata %>%
  filter(lakename %in% c("Paul Lake", "Peter Lake")) # %in% = "include", c()=list

# What is another way to use the filter command to get to the same result?
NTLdata_PeterPaul <- NTLdata %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") #boolean operators, | = "or"

# Make two data frames, one for each lake
Pauldata <- NTLdata %>%
  filter(lakename == "Paul Lake")
Peterdata <- NTLdata %>%
  filter(lakename == "Peter Lake")

# How long did the monitoring last?
min(Pauldata$sampledate)
max(Pauldata$sampledate) #over 30 years of data
min(Peterdata$sampledate)
max(Peterdata$sampledate)

# Which depths are sampled in each lake?
unique(Pauldata$depth) #depth in m
unique(Peterdata$depth)
# Why didn't we use the "summary" function here?
summary(Peterdata$depth) #doesn't tell us that they sample at 0.5 increments 

# QA the one data point at 20 m depth. 
Pauldata <- Pauldata %>%
  filter(depth < 20)

```

### Exploratory data visualization

Let's make a plot of temperatures by depth. There are a lot of points, so adding a 50 % transparency to the points helps us see where points are densely clustered together.

Let's also try using the traditional limnological graph type, with depth on the y axis in reverse, to simulate a cross section of a lake. When and where do we usually observe high and low temperatures?
```{r}
ggplot(Pauldata, aes(x = depth, y = temperature_C)) + 
  geom_point(alpha = 0.5) + 
  labs(y = expression("Temperature "(degree*C)), x = "Depth (m)")
          #aes: axes, color, shape...
          #alpha = transparency

#flip axes for traditional limnology representation
ggplot(Pauldata, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", color = "DOY")
          # one sampling time in november when its really cold. 
          # thermally mixed when the temperature is the same all the way down 
          # stratified when there is a gradient in temperature within the water column (resistance to mixing, differing densities with temperature)
                  #epilimnion = top
                  #hypolimnion = bottom - water is most dense at 4C (winter)
              # dimictic - mixing twice per year (spring and fall) 


```

How do temperatures at the surface compare to temperatures at the mid-depths and at the bottom?
* temperature decreases with depth
* temperature is more variable at the upper portion of the water column
* there was one dataset where the entire water column is cold


### Graphing seasonal water profiles

Lakes in the North Temperate Lakes LTER are dimictic, meaning they mix fully twice per year. When the lakes are not mixed (stratified), the top layer (epilimnion) and the bottom layer (hypolimnion) of the lake are different temperatures.

For background information on lake stratification, see this [stratification summary article from the North American Lake Management Society](https://www.nalms.org/secchidipin/monitoring-methods/temperature-and-oxygen/).


In 1993, Paul Lake was sampled from May 20 through November 2. Let's look at temperature profiles of the lake across the year and also explore how this impacts dissolved oxygen.

```{r}
ggplot(subset(Pauldata, year4 == 1993), 
       aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", color = "DOY") 

# Now, make a plot as above but with dissolved oxygen saturation instead of temperature.
ggplot(subset(Pauldata, year4 == 1993), 
       aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("DO (mg/L)"), y = "Depth (m)", color = "DOY") 

#stratified conditions: DO concentrations are higher in the epilimnion (upper water column) due to wave motion (incorporating oxygen through wind towards atmospheric equilibrium), plants an algae (phytoplankton photosynthesizing during the day)
#bulge in DO at slightly lower depths (right below the thermocline) becasue phytoplankton need light and nutrients to photosynthesize. respiration occurs in the hypolimnion (low water column) giving higher nutrient concentrations in the hyperlimnion. they like to hang out where theres light from above and nutrients from below. 

```

Why does mixing occur in the spring and fall? What are the mechanisms that make this possible?
<add notes here>



### Calculating thermoclines

[rLakeAnalyzer documentation](https://cran.r-project.org/web/packages/rLakeAnalyzer/rLakeAnalyzer.pdf)

```{r}
Pauldata_thermo <- Pauldata %>%
  group_by(year4, daynum, sampledate) %>%
  # calculate thermoclines based on temperature profiles. 
  # seasonal = FALSE calculates the thermocline as the maximum density gradient
  # seasonal = TRUE calculates the thermocline as the deepest density gradient
  summarise(thermocline = thermo.depth(wtr = temperature_C, depths = depth, seasonal = FALSE)) %>%
  # remove all thermoclines within 1 m of the surface. these can represent transient stratification. warming in upper layer of epilimnion due to diurnal warming that goes away within a day or two. 
  filter(thermocline > 1)

ggplot(Pauldata_thermo, aes(x = daynum, y = thermocline)) +
  geom_point() 
  # general pattern: thermocline deepening and narrowing through the year. day 0 is day of first stratification (e.g. switch from spring to summer)
```

How does thermocline depth change over the course of the year? When do we observe the most variability in thermocline depth?
<add notes here>


### Climate change

Increases in global temperature are predicted to cause several changes to lake thermal conditions, including: 

* Increases in surface temperature
* Increases in surface minimum temperature
* Increases in extent of stratification
* Increases in length of stratification
* Decreases in ice cover (switch from dimictic to monomictic)

Several studies have addressed this topic, using long-term and spatially diverse datasets: 

https://link.springer.com/article/10.1007/s10584-015-1326-1?sa_campaign=email/event/articleAuthor/onlineFirst&error=cookies_not_supported&error=cookies_not_supported&code=2b415e25-de4c-452f-bd02-2cceae08b7a3&code=e63aabb9-76d3-4e49-b36c-e591007a9e9c

http://hpkx.cnjournals.com/uploadfile/news_images/hpkx/2020-07-15/10.1038-s43017-020-0067-5.pdf

https://agupubs.onlinelibrary.wiley.com/doi/pdf/10.1002/2015GL064097

https://link.springer.com/article/10.1007/s10584-019-02465-y

Let's explore how these changes might manifest in Paul Lake. 
```{r}
ggplot(Pauldata_thermo, aes(x = daynum, y = thermocline, color = year4)) +
  geom_point() +
  scale_color_viridis_c()
# first day of stratificaiton earlier with thermocline deepening throughout the season: first sample may be deeper earlier in the season with potential narrowing in the band earlier. 
# exercise: put the legend on top and relabel the aesthetics

ggplot(subset(Pauldata, depth == 1),
       aes(x = daynum, y = temperature_C, color = as.factor(year4))) +
  geom_point(alpha = 0.5) +
  geom_line() +
  scale_color_viridis_d()
#plotting surface temperatures. year plotted as factor (all listed individually)
# exercise: relabel the aesthetics

```

## Closing Discussion

What are the main concepts you learned about the physical properties of lakes today? What was the evidence for these concepts in the dataset?
