---
title: "3: Physical Properties of Rivers"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Analyze seasonal and interannual characteristics of stream discharge 
2. Compare discharge patterns in different regions of the United States
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What are the physical properties of streams and rivers? What variables might be of interest?

* morphology: cross sectional shape
** length, width, depth, sinuosity
* discharge, velocity
* sediment load and type
* slope

## Session Set Up

```{r, message = FALSE}
getwd()

library(tidyverse)
library(dataRetrieval)
library(cowplot)
library(lubridate)

theme_set(theme_classic())
```

## Discharge

One of the most important physical characteristics of a stream or river is **discharge**, the volume of water moving through the stream over a given amount of time. Discharge can be measured directly by measuring the velocity of flow in several spots in a stream and multiplying the flow velocity over the cross-sectional area of the stream. However, this method is effort-intensive. An easier way to approximate discharge is by developing a **rating curve** for a stream at a given sampling point. To develop a rating curve, a series of measurements of discharge and stream stage must be made together to develop a mathematical relationship. Then, subsequent discharge measurements can be calculated from a measurement of stream stage, measured by a gage that takes readings remotely. 

Need volume and velocity. Rating curve used to measure discharge using the stage. 
"readNWISdv" --> retrieve USGS data

```{r}
EnoDischarge.ratingcurve <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = c("00060", "00065"),
                     startDate = "2012-01-01",
                     endDate = "2021-01-01")

names(EnoDischarge.ratingcurve)[4:7] <- c("Discharge", "D.Approval.Code",
                                          "Gage.Height", "GH.Approval.Code")

ggplot(EnoDischarge.ratingcurve, aes(x = Discharge, y = Gage.Height)) +
  geom_point(alpha = 0.5) 

```

>Plot shows that gage height increases with greater discharge. This relationship is very tight and consistent. A regression analysis may produce an R^2 value > 95%. You can use this to estimate discharge of that site from the gage height. New rating curves are developed for each site that is gaged because stage is dependent on the specific geometries of that channel. 

>Discharge in cubic feet per meter is calculated by measuring the cross sectional area and the velocity at 0.6*depth (or other factor based on how big the river is) of the river. 

### Data import
We will be using the dataRetrieval package, which allows us to access USGS hydrologic data. We will be using the hydrologic data capabilities of this package today, but the package also comes with additional functionality to access water quality data from the Water Quality Portal. 

dataRetrieval Vignette:
https://cran.r-project.org/web/packages/dataRetrieval/vignettes/dataRetrieval.html

The function `whatNWISdata` allows us to access metadata about a site, including what variables have been monitored at that site, start and end dates, and the number of samples for a given variable.

We will be analyzing data from the following rivers/streams today: 

* Eno River in North Carolina
* Verde River in Arizona (the largest perennial stream in AZ)
* Bitterroot River in Montana
* Sauk River in Minnesota (Dr. Salk's hometown)
* Nehalem River in Oregon

```{r}
HawSummary <- whatNWISdata(siteNumbers = "02096500")
VerdeSummary <- whatNWISdata(siteNumbers = "09504000")
BitterrootSummary <- whatNWISdata(siteNumbers = "12344000")
SaukSummary <- whatNWISdata(siteNumbers = "05270500")
NehalemSummary <- whatNWISdata(siteNumbers = "14299800")
```

What data are available from the sites? Pay particular attention to the parameter codes and statistic codes.

>Daily mean, max, and min. 

What is the period of record for mean daily discharge at each of these sites?

> Eno River: 

> Verde River: 

> Bitterroot River: 

> Sauk River: 1909-2022

> Nehalem River: 

The function `readNWISdv` allows us to access daily hydrologic data, which include discharge, gage height, temperature, precipitation, and pH. The function `readNWISqw` allows us to access water quality data. 

We will access the last 10 years of data for our purposes today. Alternatively,  you could code `startDate = ""` to indicate you want to access all available data for the entire period of record. Then, you could filter out the years you don't want to analyze.

```{r}
# Import data from last 10 years
EnoDischarge <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2012-01-01",
                     endDate = "2021-01-01")

VerdeDischarge <- readNWISdv(siteNumbers = "09504000",
                     parameterCd = "00060", # discharge (ft3/s)
                    startDate = "2012-01-01",
                     endDate = "2021-01-01")

BitterrootDischarge <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2012-01-01",
                     endDate = "2021-01-01")

SaukDischarge <- readNWISdv(siteNumbers = "05270500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2012-01-01",
                     endDate = "2021-01-01")

NehalemDischarge <- readNWISdv(siteNumbers = "14299800",
                     parameterCd = "00060", # discharge (ft3/s)
                    startDate = "2012-01-01",
                     endDate = "2021-01-01")

```

### Data Wrangling
```{r}

# Renaming columns (one method of multiple - can also bind rows)
names(EnoDischarge)[4:5] <- c("Eno.Discharge", "Eno.Approval.Code")
names(VerdeDischarge)[4:5] <- c("Verde.Discharge", "Verde.Approval.Code")
names(BitterrootDischarge)[4:5] <- c("Bitterroot.Discharge", "Bitterroot.Approval.Code")
names(SaukDischarge)[4:5] <- c("Sauk.Discharge", "Sauk.Approval.Code")
names(NehalemDischarge)[4:5] <- c("Nehalem.Discharge", "Nehalem.Approval.Code")

# Remove agency_cd column from each data frame
EnoDischarge <- select(EnoDischarge, -agency_cd)
VerdeDischarge <- select(VerdeDischarge, -agency_cd)
BitterrootDischarge <- select(BitterrootDischarge, -agency_cd)
SaukDischarge <- select(SaukDischarge, -agency_cd)
NehalemDischarge <- select(NehalemDischarge, -agency_cd)

# Join data frames (in sequence)
CombinedDischarge <- full_join(EnoDischarge, VerdeDischarge, by = "Date") %>%
  full_join(., BitterrootDischarge, by = "Date") %>%
  full_join(., SaukDischarge, by = "Date") %>%
  full_join(., NehalemDischarge, by = "Date")
```

### Data Visualization
```{r}
# Build a ggplot (but remove the date column for all but the last one in order to stack them on top of each other with cowplot)
EnoPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Eno.Discharge)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"), title = "Eno River") + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
EnoPlot
#(North Carolina) flashiness from urban area, dominated by rain events, seasonal

VerdePlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Verde.Discharge)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"), title = "Verde River") + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
VerdePlot
#(Arizona)arid climate, dominated by rain events, big stream that is ephemeral

BitterrootPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Bitterroot.Discharge)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"), title = "Bitterroot River") + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
BitterrootPlot
#(West) strong annual patterns of snow melt, many tributaries, not wholy dominated by event based precipitation, high gradient (mountainous)

SaukPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Sauk.Discharge)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"), title = "Sauk River") + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())         
SaukPlot
#(Minnesota) seasonality in some years, but in later years there is a shift in seasonal pattern. Seasonality due to snow pack. Possibly less snow in 2016-18 period. Drains more low gradient system. 

NehalemPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Nehalem.Discharge)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"), title = "Nehalem River") + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12))
NehalemPlot
#(Pacific Northwest) seasonal, small watershed, affected by climate patterns but tempered by the Pacific Ocean. Urban area? Landuse impacts of deforestation. Wet and dry seasons. Winter peak instead of spring peak. Not dominated by snowpack. 

CombinedPlot <- 
  plot_grid(EnoPlot, VerdePlot, BitterrootPlot, SaukPlot, NehalemPlot, 
            ncol = 1)
CombinedPlot

# add relative heights and align axes 
CombinedPlot2 <- 
  plot_grid(EnoPlot, VerdePlot, BitterrootPlot, SaukPlot, NehalemPlot, 
            ncol = 1, rel_heights = c(1, 1, 1, 1, 1.2), align = "hv")
CombinedPlot2
```

How does the magnitude of discharge compare for these rivers?

>

Which of these rivers have a seasonal cycle of discharge? How might you explain the presence and timing of seasonal cycles based on climatic conditions?

>Bitterroot and Nahalem Rivers both have strong seasonal cycles. Verde River is an ephemeral stream located in an arid region that only flows during storm events. 

Hypothetical question: If you wanted to create a ggplot using facets instead of `plot_grid`, how would you go about wrangling your dataset differently? 

>

## Defining "typical" discharge pattern
Let's take the Bitterroot River, which displays a clear seasonal cycle. Let's define the median discharge on a given day of year and overlay this onto the range experienced. We will take a longer period of record this time.
```{r}  
BitterrootDischarge.Historical <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "") #grab whole period of reference when no start date specified

names(BitterrootDischarge.Historical)[4:5] <- c("Discharge", "Approval.Code") #rename columns to something meaningful

BitterrootDischarge.Historical <- mutate(BitterrootDischarge.Historical, 
                                         DOY = yday(Date)) #add new column for day of year

#create summary across all years
#group_by(DOY) to get typical year
#p95 and p5 to get highs and lows
#na.rm = TURE to tell it to remove any NAs
BitterrootDischarge.Pattern <- BitterrootDischarge.Historical %>%
  group_by(DOY) %>%
  summarise(Median.Discharge = median(Discharge, na.rm = TRUE), 
            p95.Discharge = quantile(Discharge, 0.95, na.rm = TRUE),
            p5.Discharge = quantile(Discharge, 0.05, na.rm = TRUE))

#put median, p95, and p5 on graph. specify y-aes separately for each one. color assigned outside of aesthetics. 
ggplot(BitterrootDischarge.Pattern, aes(x = DOY)) +
  geom_line(aes(y = Median.Discharge)) +
  geom_line(aes(y = p95.Discharge), color = "gray") +
  geom_line(aes(y = p5.Discharge), color = "gray") +  
  labs(x = "Day of Year", y = expression("Discharge (ft"^3*"/s)")) 

#peak at same time --> snowmelt!!! magnitude shift from year to year instead of timing shift. 
```


## Closing Discussion
What are your impressions of seasonal and interannual discharge for our example rivers? How might you go about interpreting discharge data for any given river?
