---
title: "6: Water Quality in Lakes"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Navigate and explore the LAGOSNE database and R package
2. Evaluate lake water quality using the trophic state index
3. Analyze spatial and temporal patterns of water quality across the northeast U.S.

## Opening Discussion
What are the major water quality impairments experienced in lakes? 
* Nutrient/fertilizer pollution that result in eutrophication (nitrogen and phosphorus)
* Sediment loading that results in higher turbidity and physically covers important habitats
* Thermal pollution that affect species distributions and communities
* Heavy metals, pesticides, cations, herbicides (water quality portal)
* Pathogens, bacteria, fungi, viruses (e.g. coliform)

## Session Set Up

```{r}
getwd()
#setwd("/Users/ataliefischer/Desktop/WDA/Water_Data_Analytics_2022")

library(tidyverse)
library(lubridate)
#install.packages("LAGOSNE")
library(LAGOSNE)

theme_set(theme_classic())
options(scipen = 100) # Tells R how much to abbreviate a number (scientific notation)

lagosne_get(dest_folder = LAGOSNE:::lagos_path(), overwrite = TRUE) # Downloads Lagos NE database onto the computer. This is a linked database with many different datasets. 

```

## Getting to know the LAGOSNE database
Navigate to https://lagoslakes.org/. We will explore this website to learn about the LAGOS-NE dataset, research, and data management and use initiatives undergone by the research team. 

LAGOS (Spanish for lakes) is a research program for the interdisciplinary, continental-scale study of lakes through time.

Useful introductions to the LAGOSNE R Package can be found here: 

https://github.com/cont-limno/LAGOSNE
https://cont-limno.github.io/LAGOSNE/articles/lagosne_structure.html How things are organized and linked together in the database.

```{r}
# Load LAGOSNE data into R session. This loads as a list of dataframes instead of a single dataframe. You will need to pull each dataframe individually. 
LAGOSdata <- lagosne_load()
names(LAGOSdata) #The big ones to know are "lakes_limno", "epi_nutr", "locus", huc codes, county and state information. 

# If the package installation and data download has not worked, use this code: 
# load(file = "./Data/Raw/LAGOSdata.rda")

# Exploring the data types that are available
LAGOSlocus <- LAGOSdata$locus # lat, long, metadata (area, perimeter), relevant zones (huc), EPA eco regions (edu_zoneid). One row per lake, with each lake having a unique identifier. 
LAGOSstate <- LAGOSdata$state # Only has 17 rows for each state. IF you want specific state data, you will need to link the datasets. 
LAGOSnutrient <- LAGOSdata$epi_nutr # Contains sample dates, with multiple rows for each lake. Chlorophyll, color, nitrogen, dissolved organic carbon, ammonium. Each nutrient has multiple columns associated with each, but the one that does not have a tag after it is the actual nutrient concentration.

# Tell R to treat lakeid as a factor, not a numeric value. These are unique identifiers, but not continuous variables. 
LAGOSlocus$lagoslakeid <- as.factor(LAGOSlocus$lagoslakeid)
LAGOSnutrient$lagoslakeid <- as.factor(LAGOSnutrient$lagoslakeid)
```

## Wrangling data frames in LAGOSNE

LAGOSNE is stored in several pieces, comprising metadata about given lakes (one observation per lake), metadata about each state (one observation per state), and data collected from lakes (one to many observations per lake over time). To connect observations from one data frame to the next, we need to find a common variable between the data frames.

For example, let's find out how many lakes are in each state. Note that LAGOSlocus only includes the state_zoneid, whereas  LAGOSstate connects state_zoneid for each state.

> These are called "star datasets" that can be linked together. 


```{r}
# Join data frames
LAGOSlocations <- left_join(LAGOSlocus, LAGOSstate, by = "state_zoneid") # Link by unique identifier. If multiple columns with the same info but different names, then use by = c("colname1" = "colname2")

ggplot(LAGOSlocations, aes(x = state)) + 
  geom_bar(stat = "count") # Count number of rows in the dataset. Can order the states by latitude or longitude! 
# Minnesota has the most lakes with information in the Lagos database

ggplot(LAGOSlocations, aes(x = state, y = lake_area_ha)) + 
  geom_boxplot() +
  scale_y_log10()

```

## Trophic state as a metric for water quality
Robert Carlson's 1977 paper, "A trophic state index for lakes," established the first numeric categorization system for a lake's **trophic state**, the amount of biomass a given system can sustain (think about nutrients and primary production). Trophic state is a useful water quality metric, as it can give insight into the propensity of a system to develop algal blooms, the degree of nutrient loading in the system, and a range of other potential water quality concerns experienced by other lakes with a similar trophic state (e.g., hypoxia). 

To calculate the **Trophic State Index**, three variables can be used. Note these should not be used to define trophic state but as indicators of the broader condition. Comparing these values in a given lake can give insight into the broader mechanisms at play.

* *chlorophyll a concentration*, a proxy for algal (phytoplankton) biomass. Pros: direct measure of primary productivity
* *Secchi disk transparency*, a measure of water clarity. Pros: simple and cheap. Cons: may yield a high TSI in highly colored lakes and in lakes where particulate matter is comprised of non-algal material (dissolved organic matter, suspended sediment)
* *Total phosphorus (TP)*, a nutrient essential for growth of primary producers. Assumptions: phosphorus is the limiting nutrient for phytoplankton growth (this assumption often holds only for summer months) (may also be that nitrogen is the actual limiting nutrient) (this can be measured in tissues, solids, and dissolved nutrients)

$TSI(Chl) = 10(6 - (2.04 - 0.68 ln Chl/ln 2))$

$TSI(SD) = 10(6 - (ln SD/ln 2))$

$TSI(TP) = 10(6 - (ln(48/TP)/ln 2))$

TSI values correspond to the following trophic states: 
**0-40**: Oligotrophic (low primary production)
**40-50**: Mesotrophic (middle PP)
**50-70**: Eutrophic (high PP) 
**70-100**: Hypereutrophic (extremely high PP)

These are not meant for management tools or to define the state of the lake overall. 

### Exploring the LAGOS nutrient data frame
```{r}
dim(LAGOSnutrient)
class(LAGOSnutrient$sampledate) # Package already reads it as a date

LAGOStrophic <- LAGOSnutrient %>%
  left_join(., LAGOSlocations, by = "lagoslakeid") %>% # One dataframe already specified with the pipe
  select(lagoslakeid, sampledate, chla, tp, secchi, 
         gnis_name, lake_area_ha, state, state_name) %>% # Selecting specific columns
  mutate(sampleyear = year(sampledate), 
         samplemonth = month(sampledate)) %>% # Create year and month column
  drop_na(chla, tp, secchi, state) # Drop any row that has NA anywhere

```

Let's look at observations of chl, secchi depth, and TP seasonally and over the period of study.

Add a line of code to facet the graphs by state.
```{r}
ggplot(LAGOStrophic, aes(x = as.factor(samplemonth), y = chla)) +
  geom_jitter(alpha = 0.2, size = 0.5) + # Gives more information about sample size and their distribution
  geom_boxplot(outlier.shape = NA, color = "#1f968b", fill = NA) +
  scale_y_log10() +
  labs(x = "Month", y = expression("Chl a ("*mu*"g/L)"))
# higher primary production in the warmer months (summer and fall). There are a lot more sampling in the summer months and way less in the winter months. 

ggplot(LAGOStrophic, aes(x = as.factor(samplemonth), y = secchi)) +
  geom_jitter(alpha = 0.2, size = 0.5) +
  geom_boxplot(outlier.shape = NA, color = "#1f968b", fill = NA) +
  scale_y_log10() +
  labs(x = "Month", y = "Secchi depth (m)") 

ggplot(LAGOStrophic, aes(x = as.factor(samplemonth), y = tp)) +
  geom_jitter(alpha = 0.2, size = 0.5) +
  geom_boxplot(outlier.shape = NA, color = "#1f968b", fill = NA) +
  scale_y_log10() +
  labs(x = "Month", y = expression("TP ("*mu*"g/L)"))
```

What do you notice about the seasonality of these variables? If we were to characterize a lake based on the value of a given variable, how might seasonality affect our interpretations?

>

### Calculating trophic state index

Let's add a TSI value calculated from each of the three variables to the data frame. Let's also add a column that designates the lake as oligotrophic, mesotrophic, eutrophic, or hypereutrophic based on the TSI.chl value. 

```{r}
LAGOStrophic <- LAGOStrophic %>%
  filter(samplemonth %in% c(5:9)) %>% # Capturing summer period
  mutate(TSI.chl = round(10*(6 - (2.04 - 0.68*log(chla)/log(2)))),
         TSI.secchi = round(10*(6 - (log(secchi)/log(2)))), 
         TSI.tp = round(10*(6 - (log(48/tp)/log(2)))), 
         trophic.class = case_when(TSI.chl < 40 ~ "Oligotrophic",
                                   TSI.chl >= 40 & TSI.chl < 50 ~ "Mesotrophic",
                                   TSI.chl >= 50 & TSI.chl< 70 ~ "Eutrophic",
                                   TSI.chl >= 70 ~ "Hypereutrophic")) # Similar to ifelse function

class(LAGOStrophic$trophic.class)

LAGOStrophic$trophic.class <- factor(LAGOStrophic$trophic.class,
                                     levels = c("Oligotrophic", "Mesotrophic", "Eutrophic", "Hypereutrophic"))

#scales::show_col(colormap(colormap = colormaps$magma, nshades=16))

# Let's save the LAGOStrophic dataset as a csv in our Data folder. 

write.csv(LAGOStrophic, file = "./Data/Processed/LAGOStrophic.csv", row.names = FALSE)
```

Now let's compare TSI values calculated from chl, secchi depth, and TP. If these were all perfectly equivalent metrics, all points should line up on the 1:1 line. 

```{r}

ggplot(LAGOStrophic, aes(x = TSI.chl, y = TSI.secchi, color = samplemonth)) +
  geom_point(alpha = 0.1) + 
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(limits = c(0, 100)) +
  geom_abline(slope = 1, intercept = 0) +
  scale_color_viridis_c(option = "magma", end = 0.8, direction = -1) +
  theme(legend.position = "top")

ggplot(LAGOStrophic, aes(x = TSI.chl, y = TSI.tp, color = samplemonth)) +
  geom_point(alpha = 0.1) + 
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(limits = c(0, 100)) +
  geom_abline(slope = 1, intercept = 0) +
  scale_color_viridis_c(option = "magma", end = 0.8, direction = -1) +
  theme(legend.position = "top")

```

What is the observed relationship, and how does it depart from the 1:1 relationship? What do values above or below the 1:1 line tell us about the conditions present in a lake? (see NALMS description)[https://www.nalms.org/secchidipin/monitoring-methods/trophic-state-equations/]

> 

### Comparing TSI across states

Here are two different ways to visualize TSI across the states in the LAGOS-NE database. 

Exercise: create a data frame summarizing the count of each trophic class by state. Then, recreate the bar plot but with relative proportions of lakes in each trophic class (adding up to 1, stacked on top of each other) rather than counts.

```{r}
ggplot(LAGOStrophic, aes(x = state, y = TSI.chl)) +
  geom_violin(draw_quantiles = 0.50) +
  geom_hline(yintercept = 40, lty = 2, color = "gray") +
  geom_hline(yintercept = 50, lty = 2, color = "gray") +
  geom_hline(yintercept = 70, lty = 2, color = "gray")

ggplot(LAGOStrophic, aes(x = state, fill = trophic.class)) +
  geom_bar(stat = "count", position = position_dodge(preserve = "single")) +
  theme(legend.position = "top") +
  labs(x = "", y = "Number of Observations", fill = "Trophic Class") +
  scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8, direction = -1)

```

## Closing Discussion
What factors might you expect to influence trophic state in a given lake? What are some variables in the LAGOSNE database that could help you test your hypothesis?
