---
title: "Assignment 8: Time Series Analysis"
author: "Atalie Fischer"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on time series analysis

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/dKEutwXiFewkSTwN9

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-03-29

## Setup

1. Verify your working directory is set to the R project file. Load the tidyverse, lubridate, trend, forecast, and dataRetrieval packages. Set your ggplot theme (can be theme_classic or something else).


```{r setup}
getwd()

library(tidyverse)
library(lubridate)
library(trend)
library(forecast)
library(dataRetrieval)

theme_set(theme_classic())
```

## Data Import and Processing

2. Import discharge data (parameter 00060) from Clear Creek, Colorado (site 06719505) from the start of 1975 through the end of 2021.
3. Graph Clear Creek discharge over time. 
4. Create a new data frame with the sum of monthly discharge in acre-feet per month.

```{r}
#import discharge data from Clear Creek, Colorado
ClearCreek_flow <- readNWISdv(siteNumbers = "06719505",
                     parameterCd = "00060", # discharge (cfs)
                     startDate = "1976-01-01",
                     endDate = "2021-12-31")
names(ClearCreek_flow)[4:5] <- c("Discharge", "Approval.Code")

#graph Clear Creek discharge over time
ggplot(ClearCreek_flow, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(title = "Discharge of Clear Creek, Colorado")

# generate total monthly discharge
ClearCreek_flow_monthly <- ClearCreek_flow %>%
  mutate(Year = year(Date),
         Month = month(Date)) %>%
  group_by(Year, Month) %>%
  summarise(Discharge_acftmo = sum(Discharge)*1.98347) #convert to acre-feet/month


```


## Time Series Decomposition
5. Create a time series of discharge from the monthly data frame. Make sure to add start and end dates like we did in class.
6. Decompose the time series using the `stl` function.
7. Visualize the decomposed time series.

```{r}
ClearCreek_flow_monthly_ts <- ts(ClearCreek_flow_monthly$Discharge_acftmo, frequency = 12,
                           start = c(1976, 1, 1), end = c(2021, 12, 1))
ClearCreek_flow_monthly_decomp <- stl(ClearCreek_flow_monthly_ts, s.window = "periodic")
plot(ClearCreek_flow_monthly_decomp)
```

8. How do the seasonal and trend components of the decomposition compare to the Neuse River discharge dataset? 

> Seasonal: Clear Creek shows a much stronger seasonal cycle compared to the Neuse River, likely due to the domination of snowmelt in the system.

> Trend: The trend of Clear Creek and the Neuse River are pretty similar, both showing increases and decreases over time. 

## Trend Analysis
Research question: Has there been a monotonic trend in discharge in Clear Creek over the period of study?

9. Run a Seasonal Mann-Kendall test on the monthly discharge data. Inspect the overall trend and the monthly trends. 

```{r}
ClearCreek_flow_monthly_trend <- smk.test(ClearCreek_flow_monthly_ts)

ClearCreek_flow_monthly_trend
summary(ClearCreek_flow_monthly_trend)
```

10. Is there an overall monotonic trend in discharge over time? Are there monthly trends over time? If so, are they positive or negative?

> There is no overall monotonic trend in discharge over time, with a p-value of 0.8354 (< alpha = 0.05). There is no significant positive or negative monthly trends over time. 


## Forecasting
Research question: can we predict discharge in Clear Creek moving into the future?

11. Run the auto.arima function on the Clear Creek time series to search for the best fit. Create an object that defines the best fit model. 
12. Make a prediction into the future and plot the future predictions. 

```{r}
auto.arima(ClearCreek_flow_monthly_ts, trace = TRUE)
fit <- arima(ClearCreek_flow_monthly_ts, c(1,0,1),seasonal = list(order = c(2,1,0), period = 12))
ClearCreekprediction <- predict(fit, n.ahead = 10*12)
ts.plot(ClearCreek_flow_monthly_ts, ClearCreekprediction$pred, lty = c(1, 3))
```

13. How did the forecasting for Clear Creek compare to the Neuse River?

> The forecasting for Clear Creek is much more reasonable than for the Neuse River, showing seasonal variation in discharge due to snowmelt. However, variability is not shown. 